---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Tegan Lochner"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

In this analysis, I will look at median gross rent in Illinois. I will explore which communities are measured the most and least reliably, and what impacts that could have on public policy. 

---

## Choosing what to measure


```{r}
my_state <- "IL"        # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**[There are 102 rows of data which correlates to the 102 counties in Illinois. ]**


---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data |> 
  mutate(moe_pct = moe / estimate * 100)
```


Now sort to see which counties are measured least reliably.

```{r}
county_data |> 
  arrange(desc(moe_pct)) |> 
  select(NAME, estimate, moe, moe_pct) |> 
  head(10)
```


**[Hardin County, Illinois has the largest margin or error relative to its estimate. In the census data from 2019-2023, its' population is 3623, and it's margin of error percentage is 45.6%.]**


---

## Sorting counties into reliability categories

```{r}
county_data <- county_data |> 
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


**[High confidence: 38 counties, Moderate confidence: 43 counties, Low confidence: 21 counties]**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide |> 
  mutate(moe_pct = myvarM / myvarE * 100) |> 
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide |> 
  group_by(reliability) |> 
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```


**[The average population of a "high confidence" county is about 299458, a "moderate confidence" county is about 22877, and a "low confidence" county is about 7787. This indicates that counties with lower average populations tend to have less reliable, or uncertain, data regarding median gross rent.]**

---

## What this means for policy

1. **Which communities in your state do we know the least about?** 
   The counties in Illinois that we know the least about are Alexander, Brown, Gallatin, Hamilton, Hardin, Johnson, Massac, Menard, Pope, and Schuyler. All 10 of these counties are classified as "low confidence" reliability due to their large percentage of margin of error. 
   
2. **Why?** 
   The American Community Survey (ACS) is based on data that is sampled from the population, rather than surveying every single person like the Decennial Census. Counties with smaller populations will produce a smaller sample size, thus likely resulting in more uncertainty than a large county with a large sample size. When analyzing statistics, a smaller sample size will almost always have a larger margin of error than a larger sample size. This uncertainty is what causes these small counties to have "low confidence" reliability. 
   
3. **What could go wrong?** 
   If a state agency were to allocate resources using this variable, they may over-allocate or under-allocate to any of the 10 "low confidence" counties. While over-allocating may waste limited resources and taxpayer dollars, under-allocating can leave many vulnerable people in need. Conversely, it is likely that their allocations for larger counties, which typically fall in the "high confidence" reliability due to their small margins of error, will be accurate. Treating every county's number as equally certain is inherently unfair to small counties. 


**[Upon learning how the ACS operates, I was naively shocked to find out that the data is extrapolated from a much smaller sample size, not the entire population. In my childhood, I imagined government workers going door to door with detailed paper surveys that they would then spend months logging into a large complex computer software. After Friday's class, I found myself questioning the ethics of creating policy, even if it is positive, based off of data that may not be entirely comprehensive or representative of the people which whom it will affect. After completing this assignment, I am even further convinced that there must be another way to understand people, where they live, and how they live. I think that this is where I see community engagement operating most effectively. Bridging the gap between what is "data" and what is "reality". Although data analytics can be excellent at analyzing large overarching trends, I think in some ways it fails to get as fine tuned as one would hope, at least regarding census data. As spatial analysts, we have a responsibility to understand and communicate not just the great potential of data analysis, but also it's limits.]**


